---
title: "Parse ecoPCR"
output: html_notebook
---

This notebook parses reference databases for 12SV5 and trnL from Schneider et al.

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r}
library(Biostrings)
library(DECIPHER) # For Disambiguate()
library(here)
library(readxl)
library(ShortRead) # For clean()
library(tidyverse)
```

## Read in data

### 12SV5

Called "Vert01" by Schneider et al. 

```{r}
ref.raw <- 
     here('data', 
          'inputs', 
          'other-dbs', 
          'Schneider_ScientificReports_2021_SUPP.xlsx') %>% 
     read_xlsx(sheet = 'Data S1') # Vert01

names(ref.raw) <- 'line'

head(ref.raw)
```
Interesting: two sequence lines per header.  Think that these have just been text wrapped. Also interesting is that many of these have degenerate nucleotides. Will I need to specify those out?  Are these degenerate nucleotides also in the EMBL reference? 

```{r}
dim(ref.raw)
```

## Tidy
```{r}
# Header lines
headers.i <- grep('^>', ref.raw$line)
length(headers.i)

# Sequence lines
seqs.i <- grep('^[^>]', ref.raw$line)
length(seqs.i)
```
Reported # of sequences in reference from paper is 16,292. Yay!

Most sequences appear to be over two lines, but some have more.
```{r}
32595/16292
```

We want headers in one column, and sequences following them in another.
```{r}
# Helper function to join sequences from two adjacent lines
concat_split <- function(lines, seqs.i){
     # Create a new data structure
     seqs.concat <- character()
     
     # Loop over indices, concatenating sequences to one another if they occur
     # on adjacent lines
     
     current.seq = lines[seqs.i[1]] # Initialize with first sequence
     for (i in 2:length(seqs.i)){
          if (seqs.i[i] - seqs.i[i-1] == 1){ # If the seq index is +1 from prior
               current.seq = paste0(current.seq, lines[seqs.i[i]]) # Join them
          } else { 
               seqs.concat <- append(seqs.concat, current.seq) # Save prior one
               current.seq = lines[seqs.i[i]] # Start a new sequence
          }
     }
     
     # Need to append the final sequence (won't trigger else condition above)
     seqs.concat <- append(seqs.concat, current.seq)
     
     seqs.concat
}
```

```{r}
seqs <- concat_split(ref.raw$line, seqs.i)
length(seqs)
```

```{r}
hist(nchar(seqs))
```

```{r}
# Make new dataframe (can't do this until synchronized)
ref <- data.frame(header = ref.raw$line[headers.i],
                  seq = seqs)
```

Now, work on header column.  There are a variable number of entries, which makes splitting difficult.  Start by just isolating the ones I'm interested in, which are:

- accession
- taxid
- scientific_name: Am I correct in thinking that this pulls the lowest taxonomic level of those available?

- Maybe also merged_taxid? But I don't know exactly what this is.  Idea would be to dive into source code and try to figure out

```{r}
ex.acc <- '(?<=^>)\\S+' # First word following > at start of string
ex.taxid <- '(?<= taxid=)\\d+' # Digits following 'taxid='
ex.name <- '(?<= scientific_name=)[^;]+'

ref <- mutate(ref,
              accession = str_extract(header, ex.acc),
              taxid = str_extract(header, ex.taxid),
              scientific_name = str_extract(header, ex.name))

dim(ref)
```

## Subset to dietary animals

```{r}
# List of consumed animals from our curated record
# Note-- suspect fish are underrepresented here
foods <- 
     here('data', 
          'inputs',
          'human-foods.csv') %>%
     read.csv(stringsAsFactors = FALSE) %>%
     filter(category == 'animal') %>% 
     # Add human so we can filter out downstream
     bind_rows(data.frame(scientific_name = 'Homo sapiens',
                          common_name = 'human',
                          category = 'host',
                          added_by = NA))
```

```{r}
cat('Sequences from a curated food species:',
    sum(ref$scientific_name %in% foods$scientific_name), '\n')
cat('Percentage of sequences from a curated food species:', 
    sum(ref$scientific_name %in% foods$scientific_name)/dim(ref)[1], '%')
```

```{r}
ref <- filter(ref, scientific_name %in% foods$scientific_name)
```

### Sequence ambiguities

#### Preliminary: Keep only specified sequences

```{r}
# Accept only sequences with A, C, G, T nucleotides
dim(ref)
ref.known <- filter(ref, grepl('^[agct]+$', seq))
dim(ref.known)
```

```{r}
# Confirm
ref.known$seq %>% 
     DNAStringSet() %>% 
     alphabetFrequency() %>% 
     colSums()
```

```{r}
# Format as reference and save
ref.fasta <- DNAStringSet(ref$seq)
names(ref.fasta) <- paste(ref$accession, ref$scientific_name)

# writeXStringSet(ref.fasta, here('data', 'processed',
#                                 'dada2-compatible','12SV5',
#                                 '12SV5_Schneider_foods.fasta'))
```

## Taxonomy 

