---
title: "Taxa names"
output: html_notebook
---

## Setup
```{r setup}
require('knitr')
knitr::opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r}
library(here)
library(tidyverse)
```

## Read in data

### Crop Origins

```{r}
milla <- 
     here('data',
          'processed',
          'Crop_Origins_Phylo',
          'Crop_Origins_Phylo_v_live',
          'crop_origins_v_live',
          'crop_origins_live_db.csv') %>% 
     read_delim(delim = ';')

milla
```

### Foods list

```{r}
foods <- 
     here('data',
          'raw',
          'human-foods.csv') %>% 
     read_csv()

foods
```

### Current reference

```{r}
ref <- 
     here('data',
          'processed',
          'dada2-compatible',
          'trnL',
          'trnLGH.fasta') %>% 
     Biostrings::readDNAStringSet()

ref
```

## Pre-process

### Crop Origins

```{r}
milla <- 
     milla %>% 
     select(taxon = species_name,
            common_name = common_name_crop) %>% 
     mutate(common_name = gsub(pattern = '_',
                               replacement = ' ',
                               common_name))
```

### Foods list

```{r}
# Prune to plants only
dim(foods)
foods <- 
     foods %>% 
     filter(category == 'plant') %>% 
     select(taxon = scientific_name,
            common_name)
dim(foods)
```

### Reference

```{r}
# Format as dataframe
ref <- as.character(ref)

ref <- 
     data.frame(
         id = names(ref),
         seq = ref
     ) %>% 
     # Separate ID into accession and taxon
     separate(col = id,
              into = c('accession',
                       'taxon'),
              sep = ' ',
              extra = 'merge')

ref
```

## Assign ASV names

```{r}
# How many total ASVs are there?
nrow(ref)
```

```{r}
# How many distinct sequences
n_distinct(ref$seq)
```

So the goal here is to aggregate the names.

### Add common names
```{r}
# From our annotation
ref <- left_join(ref, foods)
```

```{r}
# Many are still missing:
table(is.na(ref$common_name))
```

```{r}
# How many of these turn up in Milla?
missing <- filter(ref, is.na(common_name))

missing <- left_join(missing, milla, by = c('taxon'))
table(is.na(missing$common_name.y))
```

I think the remainder are subspecies (from a species-level query) or species (from a genus-level query). May need to go in and refine these.

```{r}
ref
```

### Group

#### LCA

##### Get taxonomy

This could theoretically be redone with the MButils::asv_to_taxonomy() function, but here I also want to pipe forward the common name column, and the code doesn't yet accommodate that. 

```{r}
ref <- rename(ref, 
              index = accession,
              label = taxon,
              asv = seq)
```


```{r}
taxmap <- metacoder::lookup_tax_data(ref,
                                     type = 'seq_id',
                                     column = 'index')
```

```{r}
# Get taxonomy and join to reference data
taxonomy <- metacoder::taxonomy_table(taxmap,
                                      use_ranks = c('superkingdom',
                                                    'kingdom',
                                                    'phylum',
                                                    'order',
                                                    'family',
                                                    'genus',
                                                    'species',
                                                    'subspecies',
                                                    'varietas',
                                                    'forma'),
                                      add_id_col = TRUE)

# ID taxonomy
taxonomy.ids <- metacoder::taxonomy_table(taxmap,
                                          value = 'taxon_ids',
                                          use_ranks = c('superkingdom',
                                                        'kingdom',
                                                        'phylum',
                                                        'order',
                                                        'family',
                                                        'genus',
                                                        'species',
                                                        'subspecies',
                                                        'varietas',
                                                        'forma'))
```

```{r}
taxtab <-
     full_join(taxmap$data$query_data,
               taxonomy)
```

There will be missing entries here if they are nested (e.g. Cucumis sativus var. hardwickii will have an entry and C. sativus will not). Fill these in using the ID-based taxonomy:

```{r}
taxtab.add <-
     taxtab %>%
     filter(is.na(superkingdom) & taxon_id != 'unknown') %>%
     select(taxon_id, index, label, asv, common_name)

missing <- unique(taxtab.add$taxon_id) # Duplicates if multiple accessions
```

```{r}
# Find the row of each missing taxon in the ID-based taxonomy table
# Helper function to get first index (all matches should be identical):
first_index <- function(m){
     # m is a missing taxon ID
     which(taxonomy.ids == m, arr.ind = TRUE)[1, ]
}

rows <-
     lapply(missing, first_index) %>%
     bind_rows() %>%
     mutate(taxon_id = missing) %>%
     mutate(pad_na = ncol(taxonomy.ids) - col)
```

```{r}
# Now, extract each row, pad with NAs, and build out taxonomy entries to add
taxonomy.add <- data.frame()
# Remove taxon ID to synchronize indices now that join has been done
taxonomy <- taxonomy[, 2:ncol(taxonomy)]
for (i in seq(nrow(rows))){
     entry <- taxonomy[rows$row[i], 1:rows$col[i]] # Extract
     entry <-
          entry %>%
          as.character() %>%
          c(rep(NA, rows$pad_na[i])) # Pad
     
     names(entry) <- names(taxonomy) # Name
     
     taxonomy.add <- bind_rows(taxonomy.add, entry) # Join
}
```

```{r}
# Join back to ASVs by lowest-level label
taxtab.add <-
     taxonomy.add %>%
     MButils::lowest_level() %>%
     left_join(taxtab.add, ., by = c('label' = 'name'))
```

```{r}
# Make full taxonomy table
taxtab <-
     taxtab %>%
     filter(!(taxon_id %in% taxtab.add$taxon_id)) %>%  # Remove entries we looked up
     bind_rows(taxtab.add) %>% # Rejoin
     select(asv, superkingdom:forma)
```

##### Calculate LCA
```{r}
# Now, calculate LCA
taxtab <- 
     taxtab %>% 
     MButils::lca() %>%
     MButils::lowest_level() # Explicitly add column for LCA name
```

#### Summarize for reference
 
```{r}
taxtab %>% 
     rownames_to_column(var = 'asv') %>% 
     left_join(select(ref, asv, label, common_name)) %>% # Add encompassed food details
     group_by(asv, name) %>% 
     summarize(
          taxon = paste(label, collapse=", "),
               commmon_name = paste(common_name, collapse=", ")) %>% 
     write_csv(here('data',
                    'processed',
                    'dada2-compatible',
                    'trnL',
                    'trnLGH ASV common names.csv'))
```

Can then go on to manually edit these?